GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis

### 2023-10-11 figs/GSE158422_sensitivity.Rmd
gsa sensitivity

dots <- lapply(ssres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0)
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h= ssres[[1]][1,1])

dots <- lapply(ssres,function(x) { x[,"FP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(TPdots/FPdots)
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})

boxplot(ratio,names=SAMPLESIZES,col="white",ylab="TP per FP",xlab="sample size",cex=0)
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

Individual probes have a distribution that leans positive, but after aggregation to genes it is actually
negative.

@@

gsaup <- gsaup[order(gsaup$P.DE),]
head(gsaup,10) %>% kbl(caption="Top hypermethylated pathways") %>% kable_paper("hover", full_width = F)

gsadn <- gsadn[order(gsadn$P.DE),]
head(gsadn,10) %>% kbl(caption="Top hypomethylated pathways") %>% kable_paper("hover", full_width = F)

@@

# GMEA: promoter probes only

## Limma-aggregate-1-sample t-test promoter only

```{r,ls_promoter}

dmp <- dm[grep("Promo",dm$Regulatory_Feature_Group),]

#tic()
#dmpagg <- agg(dmp,cores=CORES)
#toc()

#head(dmpagg$gme_res_df,20) %>%
#  kbl(caption="top gmea genes (t-test)") %>%
#  kable_paper("hover", full_width = F)

>>>>

dots <- lapply(ssres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GMEA") ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h= ssres[[1]][1,1])

dots <- lapply(ssres,function(x) { x[,"FP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0) ; grid()
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(ssres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(ssres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@

Run GSA meth on the full dataset.

@@

gsaup <- gsaup[order(gsaup$P.DE),]
gsadn <- gsadn[order(gsadn$P.DE),]

gsig_up <-  subset(gsaup,FDR<0.05)
gsig_up %>% kbl(caption="Hypermethylated pathways") %>% kable_paper("hover", full_width = F)
gsig_dn <-  subset(gsadn,FDR<0.05)
gsig_dn %>% kbl(caption="Hypomethylated pathways") %>% kable_paper("hover", full_width = F)
gsig <- rbind(gsig_up,gsig_dn)

Only 4 hyper and 1 hypomethylated gene set were identified.
This is surprising as there were 1299 hypermethylated probes corresponding to 808 genes, and
1016 hypomethylated probes corresponding to 657 genes.

## GSAmeth Downsample

```{r,gsamethdownsample}

gsres <- lapply(SAMPLESIZES,function(n) {
  SEEDS <- seq(100,5000,100)
  dsres <- mclapply(SEEDS, function(SEED) {
    set.seed(SEED)
    mysample <- sample(1:37,n,replace=FALSE)*2
    mysample <- c(mysample,(mysample-1))
    mysample <- mysample[order(mysample)]
    design2 <- design[mysample,]
    design2 <- design2[,colSums(design2)>0]
    #mval downsample
    mval2 <- mval[,mysample]
    dm2 <- runlimma(mval2,design2,myann)
    sigup2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC>0))
    sigdn2 <- rownames(subset(dm2,adj.P.Val<0.05 & logFC<0))
    #gsamet
    gsaup2 <- gsameth(sig.cpg=sigup2, all.cpg=rownames(dm2), collection=gs_entrez)
    gsadn2 <- gsameth(sig.cpg=sigdn2, all.cpg=rownames(dm2), collection=gs_entrez)
    gsig_up2 <- subset(gsaup2,FDR<0.05)
    gsig_dn2 <- subset(gsadn2,FDR<0.05)
    gsig2 <- rbind(gsig_up2,gsig_dn2)
    TOT=nrow(gsig2)
    TP=length(which(rownames(gsig2) %in% rownames(gsig)))
    FP=length(which(!rownames(gsig2) %in% rownames(gsig)))
    TPES=0
    RES=c("TOT"=TOT,"TP"=TP,"FP"=FP,"TPES"=TPES)
    RES
  },mc.cores=8)
  do.call(rbind,dsres)
})

mx <- matrix(rep(0,200),ncol=4)
colnames(mx) <- c("TOT","TP","FP","TPES")

# fix empty matrix
gsres <- lapply(gsres,function(x) {
  if ( typeof(x) == "character" ) {
    mx
  } else {
    x
  }
} )

dots <- lapply(gsres,function(x) { x[,"TP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. true positives",xlab="sample size",cex=0,main="GSAmeth")
beeswarm(dots,add=TRUE,pch=1,cex=0.6)
abline(h= ssres[[1]][1,1])

dots <- lapply(gsres,function(x) { x[,"FP"] })
boxplot(dots,names=SAMPLESIZES,col="white",ylab="no. false positives",xlab="sample size",cex=0)
beeswarm(dots,add=TRUE,pch=1,cex=0.6)

dots <- lapply(gsres,function(x) { x[,"TP"] })
TPdots <- do.call(rbind,dots)
dots <- lapply(gsres,function(x) { x[,"FP"] })
FPdots <- do.call(rbind,dots)
ratio <- t(FPdots / (TPdots + FPdots))
ratio <- lapply(1:ncol(ratio),function(i) {ratio[,i]})
boxplot(ratio,names=SAMPLESIZES,col="white",ylab="proportion FP",xlab="sample size",cex=0) ; grid()
beeswarm(ratio,add=TRUE,pch=1,cex=0.6)

@@
